# Getting started with Bioconductor ecosystem

```{r echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

::: {.callout-note icon='true'}

## Aims

- Importing genomic files (coverage tracks and features) into R
- Discovering the `Rle` / `RleList` representation of genome-wide coverage
- Manipulating `GRanges` objects (resizing, centering, filtering)
- Extracting and aggregating a coverage signal over a set of features
- Prototyping the first three functions of the package we will build this week:
  `importFiles()`, `filterGRanges()` and `computeCoverage()`

:::

::: {.callout-tip icon='true'}

**At any time, if you are lost or do not understand how functions in the**
**proposed solution work, type `?<function>` in the R console **
**and a help menu will appear.**

You can also check the help tab in the corresponding quadrant.

:::

::: {.callout-tip icon='true'}

## Where we are heading

Over the week, we will build a package which can **plot the aggregated coverage**
**of a genomic track over a set of `GRanges` of interest** (at a fixed width).

Today, we prototype the core steps *interactively*, in a plain R script.
Over the next days, we will progressively turn this prototype into a proper,
documented, tested `Bioconductor` package. By the end of today, you will have
written draft versions of three of the four functions that make up the package:

```{r}
importFiles(bw_file, features_file, width) |>
    filterGRanges() |>
    computeCoverage() |>
    plotCoverage()
```

:::

## Set up your working environment

Let's create a project for this course!
In RStudio: `File` > `New Project...`, create a project entitled "Bioc-workshop".
Open the newly created project.
Download files required for the workshop from the following Google Drive Shared folder:

```
https://drive.google.com/drive/folders/1JHMSMYg8MbA9m53mxjHtcBa4_qycgaqy?usp=sharing
```

**Unzip the archive in the same directory you have created your project in.**

For today, we will work with two files stored in `Share/Day1/`:

- `Scc1-vs-input.bw`: a **coverage track** (`bigwig` format) of a cohesin (Scc1)
  ChIP-seq experiment in yeast (*S. cerevisiae*);
- `Scc1-peaks.narrowPeak`: the set of Scc1 **binding sites** (genomic features),
  as called from the same experiment.

These are the very same files we will later ship inside our package, so getting
comfortable with them now will pay off all week.

## Importing a coverage track in R

A coverage track stores, for each position of the genome, how much signal
(e.g. how many sequencing reads) was observed there. `bigwig` is the standard
binary format to store such tracks.

::: {.callout-question .icon .callout-note}

Find the appropriate `Bioconductor` function to import a `bigwig` file in R.
Import the coverage track as an `RleList` object (`as = "RleList"`).

::: {.callout-hint .icon .callout-note}

## Hint

The `rtracklayer` package provides a generic `import()` function that recognises
many genomic file formats. Its `as =` argument controls the class of the
returned object.

```{r}
library(rtracklayer)
bw_file <- here::here("Share/Day1/Scc1-vs-input.bw")
coverage <- import(bw_file, as = ______)
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
library(rtracklayer)
bw_file <- here::here("Share/Day1/Scc1-vs-input.bw")
coverage <- import(bw_file, as = "RleList")
coverage
```

:::

:::

::: {.callout-question .icon .callout-note}

What kind of object is a `RleList`? How many elements does it contain, and what
do they correspond to? What does the `Rle` ("Run-Length Encoding") representation
buy us here?

::: {.callout-hint .icon .callout-note}

## Hint

An `RleList` is a *list of `Rle` vectors*. Try `class(coverage)`,
`length(coverage)`, `names(coverage)` and inspect one element with `coverage[[1]]`.
A genome is mostly made of long stretches of identical values — think about how
run-length encoding stores `0 0 0 0 0 3 3 3` compactly.

```{r}
class(coverage)
length(coverage)
names(coverage)
coverage[[1]]
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
class(coverage)
## One Rle per chromosome
length(coverage)
names(coverage)
## Each Rle stores runs of identical coverage values, instead of one value
## per base pair. This is extremely memory-efficient for genomic coverage,
## which is piecewise-constant.
coverage[[1]]
lengths(coverage)  ## the length of each Rle = the length of each chromosome
```

:::

:::

## Importing genomic features in R

Genomic features (peaks, genes, regulatory elements, ...) are typically stored
as `bed`-like files. A `narrowPeak` file is a `bed` variant used for ChIP-seq
peaks.

::: {.callout-question .icon .callout-note}

Import the `Scc1-peaks.narrowPeak` file. What class of object do you obtain?

::: {.callout-hint .icon .callout-note}

## Hint

The same `rtracklayer::import()` function handles `bed`/`narrowPeak` files; here
you do **not** need the `as =` argument. The result is a `GRanges`.

```{r}
features <- import(here::here("Share/Day1/______"))
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
features <- import(here::here("Share/Day1/Scc1-peaks.narrowPeak"))
features
class(features)
```

:::

:::

::: {.callout-question .icon .callout-note}

How many features are there? On which chromosomes do they sit? What is their
median width?

::: {.callout-hint .icon .callout-note}

## Hint

`GRanges` behave a lot like vectors. Useful accessors: `length()`,
`seqnames()`, `width()`. Wrap `seqnames()` in `table()` to count features per
chromosome.

```{r}
length(features)
table(seqnames(features))
median(width(features))
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
length(features)
table(seqnames(features))
median(width(features))
```

:::

:::

## Resizing features to a fixed width

To compare and aggregate the coverage across many features, every feature must
span the **same number of positions**. We therefore resize all features to a
fixed width, centered on the middle of each feature.

::: {.callout-question .icon .callout-note}

Resize every feature to a window of `width = 2000` bp, centered on the feature
center. Which `GenomicRanges` function does this?

::: {.callout-hint .icon .callout-note}

## Hint

Look at `?GenomicRanges::resize`. Its `fix =` argument decides which anchor
point is kept fixed while resizing: `"start"`, `"end"` or `"center"`.

```{r}
width <- 2000
resized <- resize(features, width = width, fix = ______)
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
library(GenomicRanges)
width <- 2000
resized <- resize(features, width = width, fix = "center")
resized
table(width(resized))  ## all features now span exactly 2000 bp
```

:::

:::

## Wrap it up: your first function, `importFiles()`

You now have all the pieces to import a coverage track and a set of
(fixed-width) features together. Let's package these steps into a reusable
function — the first function of our package.

::: {.callout-question .icon .callout-note}

Write a function `importFiles(bw_file, features_file, width)` that:

1. imports the coverage track as an `RleList`;
2. imports the features as a `GRanges`;
3. resizes the features to `width`, centered on their center;
4. returns both in a named `list` (`coverage` and `features`).

::: {.callout-hint .icon .callout-note}

## Hint

Start from the skeleton below and fill in the `# TODO`s with the code you just
wrote interactively. The function should simply glue those steps together.

```{r}
importFiles <- function(bw_file, features_file, width) {
    coverage <- # TODO: import bw_file as an RleList
    features <- # TODO: import features_file as a GRanges
    features <- # TODO: resize features to `width`, centered
    list(coverage = ______, features = ______)
}
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
importFiles <- function(bw_file, features_file, width) {
    coverage <- rtracklayer::import(bw_file, as = "RleList")
    features <- rtracklayer::import(features_file)
    features <- GenomicRanges::resize(features, width = width, fix = "center")
    list(coverage = coverage, features = features)
}

l <- importFiles(
    here::here("Share/Day1/Scc1-vs-input.bw"),
    here::here("Share/Day1/Scc1-peaks.narrowPeak"),
    width = 2000
)
l
```

:::

:::

## Extracting the coverage over features

We can now extract the coverage track values sitting under each feature.
Subsetting an `RleList` by a `GRanges` returns, for each feature, the `Rle` of
coverage values spanning it.

::: {.callout-question .icon .callout-note}

Extract the coverage over the resized features. What object do you get, and how
long is each of its elements?

::: {.callout-hint .icon .callout-note}

## Hint

`RleList` objects can be subset by a `GRanges`, i.e. `coverage[features]`.
The result is another `RleList`, with one element per feature.

```{r}
cov_by_feature <- l$coverage[l$features]
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
cov_by_feature <- l$coverage[l$features]
cov_by_feature
## Each element spans `width` positions (here 2000)
head(lengths(cov_by_feature))
```

:::

:::

## Filtering features: your second function, `filterGRanges()`

Some features may sit too close to a chromosome end: once resized, their window
runs *past* the chromosome boundary. Others may sit on a chromosome that is
absent from the coverage track. In both cases, extracting the coverage would
fail or return truncated profiles.

::: {.callout-question .icon .callout-note}

Before extracting coverage, we should remove such problematic features. Write a
function `filterGRanges(l)` that takes the list returned by `importFiles()` and
keeps only the features that (a) sit on a chromosome present in the coverage
track, and (b) are fully contained within their chromosome.

::: {.callout-hint .icon .callout-note}

## Hint

The length of each chromosome is `lengths(l$coverage)` (a named vector). A
feature is safe to keep if its `start()` is `>= 1` **and** its `end()` is `<=`
the length of its chromosome. Use `%in% names(l$coverage)` to check the
chromosome is covered.

```{r}
filterGRanges <- function(l) {
    features <- l$features
    chroms <- as.character(seqnames(features))
    chrom_lengths <- lengths(l$coverage)
    keep <- # TODO: on a covered chromosome AND within bounds
    l$features <- features[keep]
    l
}
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
filterGRanges <- function(l) {
    features <- l$features
    chroms <- as.character(GenomicRanges::seqnames(features))
    chrom_lengths <- lengths(l$coverage)
    on_covered_chrom <- chroms %in% names(l$coverage)
    within_bounds <- GenomicRanges::start(features) >= 1 &
        GenomicRanges::end(features) <= chrom_lengths[chroms]
    l$features <- features[on_covered_chrom & within_bounds]
    l
}

l <- filterGRanges(l)
length(l$features)  ## possibly fewer features than before
```

:::

:::

## Aggregating the coverage: your third function, `computeCoverage()`

We finally turn the per-feature coverage into a single, aggregated profile: for
each position along the window, the **mean** coverage across all features,
together with a confidence interval.

::: {.callout-question .icon .callout-note}

Reshape the extracted coverage into a `features x positions` numeric matrix.
Why a matrix, and why `byrow = TRUE`?

::: {.callout-hint .icon .callout-note}

## Hint

Unlisting the `RleList` gives one long numeric vector (feature 1's positions,
then feature 2's, ...). Folding it back into a matrix with `ncol = width` and
`byrow = TRUE` puts one feature per row and one position per column — exactly
what we need to average column by column.

```{r}
mat <- matrix(
    as.numeric(unlist(l$coverage[l$features])),
    ncol = width, byrow = TRUE
)
dim(mat)
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
mat <- matrix(
    as.numeric(unlist(l$coverage[l$features])),
    ncol = width, byrow = TRUE
)
dim(mat)  ## nrow = number of features, ncol = width
```

:::

:::

::: {.callout-question .icon .callout-note}

Features on the **minus** strand are read 3' -> 5' along the genome. To align
all profiles in the same 5' -> 3' orientation, the rows corresponding to
minus-strand features must be reversed. Do it.

::: {.callout-hint .icon .callout-note}

## Hint

Identify minus-strand features with `strand(l$features) == "-"`, then reverse
the column order of those rows only: `mat[is_minus, ] <- mat[is_minus, rev(seq_len(width))]`.

```{r}
is_minus <- as.logical(strand(l$features) == "-")
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
is_minus <- as.logical(GenomicRanges::strand(l$features) == "-")
if (any(is_minus)) {
    mat[is_minus, ] <- mat[is_minus, rev(seq_len(width)), drop = FALSE]
}
```

:::

:::

::: {.callout-question .icon .callout-note}

Now wrap everything into a function `computeCoverage(l)` that returns a
`data.frame` with one row per position and the columns `distance` (from
`-width/2` to `width/2 - 1`), `mean`, `ci_low` and `ci_high` (bounds of the 95%
confidence interval).

::: {.callout-hint .icon .callout-note}

## Hint

Per-position mean is `colMeans(mat)`. For a 95% CI, use
`1.96 * sd / sqrt(n)`, where `sd` is `apply(mat, 2, sd)` and `n = nrow(mat)`.
The `distance` column is `seq(-width/2, width/2 - 1, by = 1)`.

```{r}
computeCoverage <- function(l) {
    features <- l$features
    width <- unique(width(features))
    mat <- # TODO: build the features x positions matrix (+ flip minus strand)
    means <- # TODO: per-position mean
    ci <- # TODO: 1.96 * per-position sd / sqrt(n)
    data.frame(
        distance = seq(-width/2, width/2 - 1, by = 1),
        mean = means,
        ci_low = means - ci,
        ci_high = means + ci
    )
}
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
computeCoverage <- function(l) {
    features <- l$features
    width <- unique(GenomicRanges::width(features))
    ## features x positions matrix
    mat <- matrix(
        as.numeric(unlist(l$coverage[features])),
        ncol = width, byrow = TRUE
    )
    ## flip minus-strand features
    is_minus <- as.logical(GenomicRanges::strand(features) == "-")
    if (any(is_minus)) {
        mat[is_minus, ] <- mat[is_minus, rev(seq_len(width)), drop = FALSE]
    }
    ## per-position mean and 95% CI
    n <- nrow(mat)
    means <- colMeans(mat, na.rm = TRUE)
    ci <- 1.96 * apply(mat, 2, sd, na.rm = TRUE) / sqrt(n)
    data.frame(
        distance = seq(-width/2, width/2 - 1, by = 1),
        mean = means,
        ci_low = means - ci,
        ci_high = means + ci
    )
}

df <- computeCoverage(l)
head(df)
```

:::

:::

## A first look at the result

We now have a tidy `data.frame` describing the aggregated coverage. Let's plot
it — this preview is exactly what tomorrow's fourth function, `plotCoverage()`,
will formalize.

::: {.callout-question .icon .callout-note}

Plot `mean` (with its confidence interval ribbon) as a function of `distance`.
Do you see an enrichment of Scc1 signal around the center of its binding sites?

::: {.callout-hint .icon .callout-note}

## Hint

Use `ggplot2`, mapping `distance` to `x` and `mean` to `y`. Add a
`geom_ribbon(aes(ymin = ci_low, ymax = ci_high))` for the confidence interval
and a `geom_line()` for the mean.

```{r}
library(ggplot2)
ggplot(df, aes(x = distance, y = mean)) +
    geom_ribbon(aes(ymin = ci_low, ymax = ci_high), alpha = 0.2) +
    geom_line()
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
library(ggplot2)
ggplot(df, aes(x = distance, y = mean)) +
    geom_ribbon(aes(ymin = ci_low, ymax = ci_high), fill = "steelblue", alpha = 0.2) +
    geom_line(color = "steelblue") +
    labs(x = "Distance to center (bp)", y = "Mean coverage") +
    theme_bw()
```

:::

:::

::: {.callout-tip icon='true'}

## Recap

In a single script, you have prototyped **three of the four functions** of our
package (`importFiles()`, `filterGRanges()`, `computeCoverage()`) and sketched
the fourth (`plotCoverage()`). Tomorrow, we will turn this prototype into a
proper `Bioconductor` package: create the package skeleton, move these functions
into `R/` files, and document them.

The reference implementation of all four functions is available in
`Share/functions/`, should you want to compare with your own version.

:::
