# Kickstarting an R package

```{r echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

::: {.callout-note icon='true'}

## Aims 

- Kickstart an R package 
- Write functions
- Document functions

:::

::: {.callout-tip icon='true'}

**At any time, if you are lost or do not understand how functions in the**
**proposed solution work, type `?<function>` in the R console **
**and a help menu will appear.** 

You can also check the help tab in the corresponding quadrant. 

:::

## Proposed package functionalities {#package-functions}

In this section, we are going to create a package which aims at: 

1. Importing `.bigwig` coverage track files as well as `.bed` files of genomic coordinates
2. Filtering imported `GRanges` to make sure they all are contained within the coverage tracks 
3. Calculating the average (+/- quantiles) of coverage of the `.bigwig` file over the `GRanges` of interest
4. Plot the aggregated `.bigwig` coverage over the `GRanges` of interest 

Eventually, the functions shipped in this package should be able to turn a pair
of `.bigwig`/`.bed` files into an aggregated coverage profile — exactly the kind
of plot you already produced by hand at the end of Day 1 (mean signal, with its
confidence interval, as a function of the distance to the center of the
features).

## Starting a new package

To start developing a new package, one may want to make sure all useful 
developer resources are already set-up. 

::: {.callout-question .icon .callout-note}

Make sure to install the developer toolkit packages.

::: {.callout-hint .icon .callout-note}

## Hint

CRAN packages are installed with `remotes::install_cran()`, and `Bioconductor`
packages with `BiocManager::install()`. You will need, among others,
`devtools`, `usethis`, `available` (from CRAN) and `biocthis`, `BiocCheck`
(from Bioconductor).

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
install.packages("remotes")
remotes::install_cran(
    c(
        "available",
        "devtools",
        "rcmdcheck", 
        "knitr",
        "pkgdown",
        "RefManageR",
        "rmarkdown",
        "rstudioapi",
        "sessioninfo",
        "styler",
        "usethis",
        "gert"
    )
)
remotes::install_cran("BiocManager")
BiocManager::install(c("BiocStyle", "BiocCheck", "biocthis"))
```

:::

:::

Now the time has come to create your very first package! 
To keep a coherent cohorte, we'll create a package named `<YOURNAME>TestPackage`. 
But first, you need to make sure that this package name is available. 

::: {.callout-question .icon .callout-note}

Verify that your package name is available, with the `available` package.

::: {.callout-hint .icon .callout-note}

## Hint

The `available` package has a function named... `available()`. It checks a name
against CRAN/Bioconductor, and also flags unintended meanings.

```{r}
available::available("______")
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
available::available("JacquesTestPackage")
```

:::

:::

If this name is available, let's create your package. 

::: {.callout-question .icon .callout-note}

Which function from `devtools` can be used to create a package skeleton?
Use it to create your package.

::: {.callout-hint .icon .callout-note}

## Hint

The function is `create_package()`. It takes the *path* of the package to
create (so mind the parent directory — e.g. your home `~`).

```{r}
library(devtools)
create_package("~/______")
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
library(devtools)
setwd("~")
create_package("JacquesTestPackage")
```

:::

:::

When creating your package, `devtools`+`RStudio` combo automatically opens a new `RStudio` window, with the package's project automatically activated. Navigate to your package RStudio project window and re-open this instruction sheet. 

::: {.callout-question .icon .callout-note}

Make sure you're in the right project using the appropriate function from `rstudioapi` package.

::: {.callout-hint .icon .callout-note}

## Hint

`rstudioapi::getActiveProject()` returns the path of the currently active
RStudio project. If it is not the right one, `usethis::proj_activate()` can
switch to it.

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
usethis::proj_activate("~/JacquesTestPackage/")
rstudioapi::getActiveProject()
```

:::

:::

::: {.callout-question .icon .callout-note}

Explore the newly created folder architecture. Do you recognize all the files and directories? 

:::


The DESCRIPTION file automatically created does not exactly match Bioconductor 
requirements. You should replace it by a better-fitted DESCRIPTION file. 

::: {.callout-question .icon .callout-note}

Find the appropriate function from `biocthis` to replace the current
DESCRIPTION file by one fitting BIOCONDUCTOR's requirements.

::: {.callout-hint .icon .callout-note}

## Hint

`biocthis` functions all start with `use_bioc_*()`. Look for the one that ends
in `_description`.

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
biocthis::use_bioc_description()
```

:::

:::

::: {.callout-question .icon .callout-note}

Manually edit the important fields from your new DESCRIPTION file: `Title` & `Author`. The remaining fields will be filled out later on. 

:::

::: {.callout-question .icon .callout-note}

Now add a LICENSE file to specify under which license you wish to publish your package.

::: {.callout-hint .icon .callout-note}

## Hint

`usethis` provides one function per common license, all named `use_*_license()`
(e.g. `use_mit_license()`, `use_gpl3_license()`, ...).

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
usethis::use_mit_license()
```

:::

:::

A nice touch is to systematically provide a `README` file. 
This provides a good overview of your package when browsed e.g. in 
GitHub, and for non-R specialists, who don't necessarily know about 
`DESCRIPTION` and vignettes. And since you are developing an R package, 
let's stick with the Rmd format for your `README`.


::: {.callout-question .icon .callout-note}

What are the benefits of having a `README.Rmd` file? 

Which function from `biocthis` can create a `.Rmd` README file?

::: {.callout-hint .icon .callout-note}

## Hint

Again a `biocthis::use_bioc_*()` function — this time one that mentions
`readme` and `rmd`. A `README.Rmd` lets you embed *runnable* R code (and its
output) directly in the README.

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
biocthis::use_bioc_readme_rmd()
```

:::

:::

That being said, package submission requires that a `REAMDE.md` file only 
is present in the repository. A way to provide this is to 1) 
`devtools::build_readme()` to parse `README.Rmd` into `README.md` 
(with R chunks processed!), and 2) add `README.Rmd` to .gitignore. 

::: {.callout-question .icon .callout-note}

Parse your `README.Rmd` into `README.md` and check the differences.

::: {.callout-hint .icon .callout-note}

## Hint

`devtools::build_readme()` knits `README.Rmd` into `README.md`. Load your
package first with `devtools::load_all('.')` so that any package code used in
the README is available.

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
devtools::load_all('.')
devtools::build_readme()
rstudioapi::documentOpen("README.Rmd")
rstudioapi::documentOpen("README.md")
```

:::

:::

A bunch of extra useful commands from `bioc/usethis` can be performed at 
this stage, e.g. to add a NEWS file, a Code of Conduct, a citation file, etc... 
Beware, each command creates/modify one/several files. Pay attention to 
the comments printed in the R console!

::: {.callout-question .icon .callout-note}

Run each of the following commands one by one and make sure to fulfill each required action highlighted in the R console. 

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
biocthis::use_bioc_news_md()  ## To add a NEWS.md file
biocthis::use_bioc_coc()  ## To add a .github/CODE_OF_CONDUCT.md file
biocthis::use_bioc_support()  ## To add a .github/SUPPORT.md file
biocthis::use_bioc_issue_template()  ## To add a .github/ISSUE_TEMPLATE/issue_template.md file
biocthis::use_bioc_citation()  ## To add a inst/CITATION file
usethis::use_lifecycle_badge("Experimental")  ## To add a "Experimental" badge to your README.Rmd
devtools::build_readme()  ## To rebuild README.md
```

:::

:::

## Write functions to populate your package

To write your first package function, you can either create an `R/<name>.R`
file manually, or run `usethis::use_r(name)`.

::: {.callout-tip icon='true'}

## Reusing Day 1

In Day 1, you already prototyped three of these functions (`importFiles()`,
`filterGRanges()`, `computeCoverage()`) in a plain script. Your job today is to
**move them into the package** (one `R/<name>.R` file each), write the missing
fourth function (`plotCoverage()`), and document them all. Prototyping in a
script, then "packaging", is exactly how real development goes!

:::

Now is time to write! Refer to the functionalities (described in
[Section 0](#package-functions)) that we are aiming to provide in this package,
and start writing 4 functions which, together, can fulfill these
functionalities. Briefly, the four functions are:

1. `importFiles()`: import a coverage track and a set of genomic features (scaled to a fixed width) together in a list;
2. `filterGRanges()`: filter the imported features to only retain those fully overlapping with covered genome segments (from the coverage track);
3. `computeCoverage()`: compute the mean coverage and its confidence interval, for each position within the features' width;
4. `plotCoverage()`: plot the results with `ggplot2`.

::: {.callout-tip icon='true'}

We suggest these exact function names because the exercises for the following
days (Day 3, 4 and 5) reuse them. Using the same names will let your package
grow smoothly across the week.

:::

::: {.callout-tip icon='true'}

These functions should work in a chain: 

```{r}
importFiles(bw_path, features_path, width) |>
    filterGRanges() |>
    computeCoverage() |>
    plotCoverage()
```

:::

The only genuinely new function today is `plotCoverage()`, which draws the
aggregated profile you sketched at the end of Day 1.

::: {.callout-question .icon .callout-note}

Write a `plotCoverage(df)` function that takes the `data.frame` returned by
`computeCoverage()` and returns a `ggplot`.

::: {.callout-hint .icon .callout-note}

## Hint

Map `distance` to `x` and `mean` to `y`. Add a `geom_ribbon()` for the
confidence interval (`ymin = ci_low`, `ymax = ci_high`) and a `geom_line()` for
the mean.

```{r}
plotCoverage <- function(df) {
    ggplot2::ggplot(df, ggplot2::aes(x = distance, y = mean)) +
        # TODO: geom_ribbon() for the confidence interval
        # TODO: geom_line() for the mean
        ggplot2::labs(x = "Distance to center (bp)", y = "Mean coverage")
}
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
plotCoverage <- function(df) {
    ggplot2::ggplot(df, ggplot2::aes(x = distance, y = mean)) +
        ggplot2::geom_ribbon(
            ggplot2::aes(ymin = ci_low, ymax = ci_high),
            fill = "steelblue", alpha = 0.2
        ) +
        ggplot2::geom_line(color = "steelblue") +
        ggplot2::labs(x = "Distance to center (bp)", y = "Mean coverage") +
        ggplot2::theme_bw()
}
```

:::

:::

::: {.callout-tip icon='true'}

The complete reference implementation of the four function files is available
in `Share/functions/`. These files provide one possible implementation
fulfilling our package's requirements.

:::

## Add function documentation

::: {.callout-question .icon .callout-note}

Write roxygen tags (`@title`, `@description`, `@param`, `@return`, ...) for each
function.

::: {.callout-hint .icon .callout-note}

## Hint

roxygen comments start with `#'` and sit directly above the function. The first
line is the title, a blank `#'` line separates blocks, and each `@param`
documents one argument. Do not forget `@export` (so the function is usable) and
`@importFrom pkg fun` (to declare the functions you borrow from other packages).

```{r}
#' Import a coverage track and a set of genomic features
#'
#' Longer description of what the function does.
#'
#' @param bw_file Path to a coverage track file.
#' @param features_file Path to a features file.
#' @param width Integer. Width to resize each feature to.
#'
#' @return A named list with `coverage` and `features`.
#'
#' @importFrom rtracklayer import
#' @importFrom GenomicRanges resize
#' @export
importFiles <- function(bw_file, features_file, width) {
    # ...
}
```

:::

:::

Once you have documented each function, you can regenerate the documentation with
`document()`.

```{r}
document()
```
