# Enriching package functionalities

```{r echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

::: {.callout-note icon='true'}

## Aims

- Adding supporting data
- Documenting how the data was generated
- Adding examples
- Adding unit tests
- Checking the package

:::

::: {.callout-tip icon='true'}

**At any time, if you are lost or do not understand how functions in the**
**proposed solution work, type `?<function>` in the R console **
**and a help menu will appear.**

You can also check the help tab in the corresponding quadrant.

:::

::: {.callout-tip icon='true'}

## Reminder

We aim to create a package which can plot the aggregated coverage of a genomic
track over a set of `GRanges` of interest (at a fixed width).

By now, your package should contain the four functions (`importFiles()`,
`filterGRanges()`, `computeCoverage()`, `plotCoverage()`) prototyped in Day 1
and documented in Day 2.

:::

## Add supporting data to your package

To let your users (and yourself!) run the examples and tests, a package should
ship small example datasets. Raw data files live in `inst/extdata/`.

You can use the toy dataset files provided in `Share/data/`
(`Scc1-vs-input.bw` and `Scc1-peaks.narrowPeak`).

::: {.callout-question .icon .callout-note}

Add these two files to the `inst/extdata/` folder of your package.

::: {.callout-hint .icon .callout-note}

## Hint

`inst/extdata/` may not exist yet — create it first. You can do this from R,
or simply from your file browser.

```{r}
dir.create("inst/extdata", recursive = TRUE, showWarnings = FALSE)
file.copy(
    from = c(
        here::here("Share/data/Scc1-vs-input.bw"),
        here::here("Share/data/Scc1-peaks.narrowPeak")
    ),
    to = "inst/extdata/"
)
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
dir.create("inst/extdata", recursive = TRUE, showWarnings = FALSE)
file.copy(
    from = c(
        here::here("Share/data/Scc1-vs-input.bw"),
        here::here("Share/data/Scc1-peaks.narrowPeak")
    ),
    to = "inst/extdata/"
)
list.files("inst/extdata")
```

Once the files are in `inst/extdata/`, you can retrieve their installed path
with `system.file()`:

```{r}
system.file("extdata", "Scc1-vs-input.bw", package = "JacquesTestPackage")
```

:::

:::

`Bioconductor` requires that you **document how each raw data file was**
**generated or obtained**. This is done, less formally, in a `.R` script placed
in `inst/scripts/`.

::: {.callout-question .icon .callout-note}

Create an `inst/scripts/make-extdata.R` file, and describe there how the two
example files were produced.

::: {.callout-hint .icon .callout-note}

## Hint

The script is documentation, not something that needs to run end-to-end: a mix
of comments and (optionally) code is fine. An example is provided in
`Share/scripts/make-extdata.R`.

```{r}
dir.create("inst/scripts", recursive = TRUE, showWarnings = FALSE)
file.edit("inst/scripts/make-extdata.R")
```

:::

:::

## Add examples to your function files

Examples illustrate how your functions work. They are integrated in the manual
pages compiled with `document()`, and are often the first thing a new user
reads.

::: {.callout-question .icon .callout-note}

Add a runnable `@examples` section to each of your functions, using the example
data you just added.

::: {.callout-hint .icon .callout-note}

## Hint

Reference the installed data with `system.file()` so that the examples work on
any machine where the package is installed. A roxygen `@examples` block looks
like this:

```{r}
#' @examples
#' bw_file <- system.file("extdata", "Scc1-vs-input.bw", package = "JacquesTestPackage")
#' bed_file <- system.file("extdata", "Scc1-peaks.narrowPeak", package = "JacquesTestPackage")
#' l <- importFiles(bw_file, bed_file, width = 2000)
#' l
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

Fully worked-out function files, with `@examples` sections added, are available
in `Share/functions/`. For instance, `importFiles()` chains into the whole
pipeline:

```{r}
#' @examples
#' bw_file <- system.file("extdata", "Scc1-vs-input.bw", package = "JacquesTestPackage")
#' bed_file <- system.file("extdata", "Scc1-peaks.narrowPeak", package = "JacquesTestPackage")
#' importFiles(bw_file, bed_file, width = 2000) |>
#'     filterGRanges() |>
#'     computeCoverage() |>
#'     plotCoverage()
```

:::

:::

Once you have added examples and re-run `document()`, you can check that they all
run without error:

```{r}
document()
run_examples()
```

## Add unit tests to your package

Unit tests are the best way to make sure your package keeps working as intended
as you keep editing it. Each test runs a bit of `R` code and compares the result
to an expectation. We use the `testthat` framework.

To start using unit tests, declare it once:

```{r}
usethis::use_testthat()
```

This creates a `tests/testthat/` folder and a `tests/testthat.R` runner.

::: {.callout-question .icon .callout-note}

Create a test file for your functions.

::: {.callout-hint .icon .callout-note}

## Hint

`usethis::use_test("<name>")` creates (and opens) `tests/testthat/test-<name>.R`.
Give it a meaningful name, e.g. the function or feature under test.

```{r}
usethis::use_test("computeCoverage")
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
usethis::use_test("computeCoverage")
```

:::

:::

A test file is a series of `test_that("description", { ... })` blocks. Inside
each block, you state **expectations** with the `expect_*()` family:

- `expect_equal(x, y)` / `expect_identical(x, y)`: `x` equals `y`;
- `expect_true(x)` / `expect_false(x)`: `x` is `TRUE` / `FALSE`;
- `expect_s4_class(x, "GRanges")` / `expect_s3_class(x, "data.frame")`: `x` has the expected class;
- `expect_named(x, c("a", "b"))`: `x` has the expected names;
- `expect_error(f())` / `expect_warning(f())`: calling `f()` raises an error / warning.

::: {.callout-question .icon .callout-note}

Write a test that checks that `importFiles()` returns a list with a `coverage`
and a `features` element, and that every feature has been resized to the
requested width.

::: {.callout-hint .icon .callout-note}

## Hint

Load the example data with `system.file()`, call `importFiles()`, then state
what you *know* must be true about the result.

```{r}
test_that("importFiles() resizes features to the requested width", {
    bw_file <- system.file("extdata", "Scc1-vs-input.bw", package = "JacquesTestPackage")
    bed_file <- system.file("extdata", "Scc1-peaks.narrowPeak", package = "JacquesTestPackage")
    l <- importFiles(bw_file, bed_file, width = 2000)
    expect_named(l, c("coverage", "features"))
    expect_true(all(GenomicRanges::width(l$features) == ______))
})
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
test_that("importFiles() resizes features to the requested width", {
    bw_file <- system.file("extdata", "Scc1-vs-input.bw", package = "JacquesTestPackage")
    bed_file <- system.file("extdata", "Scc1-peaks.narrowPeak", package = "JacquesTestPackage")
    l <- importFiles(bw_file, bed_file, width = 2000)

    expect_type(l, "list")
    expect_named(l, c("coverage", "features"))
    expect_s4_class(l$features, "GRanges")
    expect_true(all(GenomicRanges::width(l$features) == 2000))
})
```

:::

:::

::: {.callout-question .icon .callout-note}

Write a test that checks the shape and consistency of the `data.frame` returned
by `computeCoverage()`: one row per position, the expected column names, and a
confidence interval that always brackets the mean.

::: {.callout-hint .icon .callout-note}

## Hint

Run the full pipeline, then think about the invariants that must hold no matter
the data: `nrow(df) == width`, the columns are `distance`/`mean`/`ci_low`/`ci_high`,
and for every position `ci_low <= mean <= ci_high`.

```{r}
test_that("computeCoverage() returns a well-formed data.frame", {
    # ... run the pipeline into `df` ...
    expect_equal(nrow(df), width)
    expect_named(df, c("distance", "mean", "ci_low", "ci_high"))
    expect_true(all(df$ci_low <= df$mean & df$mean <= df$ci_high))
})
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
test_that("computeCoverage() returns a well-formed data.frame", {
    bw_file <- system.file("extdata", "Scc1-vs-input.bw", package = "JacquesTestPackage")
    bed_file <- system.file("extdata", "Scc1-peaks.narrowPeak", package = "JacquesTestPackage")
    width <- 2000
    df <- importFiles(bw_file, bed_file, width = width) |>
        filterGRanges() |>
        computeCoverage()

    expect_s3_class(df, "data.frame")
    expect_equal(nrow(df), width)
    expect_named(df, c("distance", "mean", "ci_low", "ci_high"))
    expect_equal(df$distance[1], -width / 2)
    expect_true(all(df$ci_low <= df$mean))
    expect_true(all(df$mean <= df$ci_high))
})
```

A complete, working example test file is available in
`Share/tests/test-computeCoverage.R`.

:::

:::

::: {.callout-tip icon='true'}

Good tests often check **edge cases** and **error handling**, not just the happy
path. For instance: does `computeCoverage()` error out cleanly if the features
do not all share the same width? You could test that with
`expect_error(...)`.

:::

Once your tests are written, run the whole suite:

```{r}
devtools::test()
```

## Check your package

Several `check` functions are available. They each check your package in a
specific way:

- `devtools::check()` verifies the basic structure of the package, the documentation, the examples and the unit tests;
- `rcmdcheck::rcmdcheck()` does somewhat the same job but builds the package first;
- `BiocCheck::BiocCheck()` checks the compliance of the package with `Bioconductor` requirements.

```{r}
devtools::check()
rcmdcheck::rcmdcheck()
BiocCheck::BiocCheck()
```

::: {.callout-question .icon .callout-note}

Attempt to fix the `ERROR`s and `WARNING`s returned by `check()` and
`BiocCheck()`.

::: {.callout-hint .icon .callout-note}

## Hint

Read each message carefully — they usually tell you exactly what to fix. Common
early offenders:

- missing `@return` / `@param` tags in the documentation;
- undeclared dependencies (add them with `usethis::use_package("<pkg>")`);
- lines longer than 80 characters (a `BiocCheck` style requirement);
- functions used without their `pkg::` prefix or an `@importFrom` tag.

Fix, re-run `document()`, then re-run the check. Iterate until it is clean.

:::

:::
