# Improving package integration

```{r echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

::: {.callout-note icon='true'}

## Aims 

- Add suppporting functions to access AnnotationHub-hosted files;
- Write a vignette to document a use case for the package;
- Create a pkgdown website for package documentation

:::

::: {.callout-tip icon='true'}

**At any time, if you are lost or do not understand how functions in the**
**proposed solution work, type `?<function>` in the R console **
**and a help menu will appear.** 

You can also check the help tab in the corresponding quadrant. 

:::

::: {.callout-tip icon='true'}

## Reminder

We aim to create a package which can plot the aggregated coverage of a genomic
track over a set of `GRanges` of interest (at a fixed width).

:::

## Use `TxDb` and `AnnotationHub` resources

### `TxDb` database

`TxDb` packages provide transcript annotations for model organisms. 

::: {.callout-question .icon .callout-note}

Find the `TxDb` package associated with UCSC-provided human gene annotations and 
install it. 

**Note:** You will need annotations from the `hg19` human genome reference.

::: {.callout-hint .icon .callout-note}

## Hint

`TxDb` packages follow the naming scheme
`TxDb.<Species>.<Provider>.<Genome>.<Table>`. For UCSC human hg19 known genes,
that is `TxDb.Hsapiens.UCSC.hg19.knownGene`. Install it with
`BiocManager::install()`.

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
BiocManager::install('TxDb.Hsapiens.UCSC.hg19.knownGene')
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene::TxDb.Hsapiens.UCSC.hg19.knownGene
```

:::

:::

`GenomicFeatures` is a package designed in order to facilitate conversion of 
`TxDb` objects into `GRanges` objects. 

::: {.callout-question .icon .callout-note}

Which function from `GenomicFeatures` can be used to find the gene loci for all genes present in the human `TxDb`?

::: {.callout-hint .icon .callout-note}

## Hint

The function is named after what it returns: `GenomicFeatures::genes()`. It
takes the `TxDb` object and returns a `GRanges` of gene loci.

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
GenomicFeatures::genes(txdb)
```

:::

:::

### `AnnotationHub` database

`AnnotationHub` provides access to a large number of NGS datasets. 

::: {.callout-question .icon .callout-note}

- Can you query the `AnnotationHub` to explore which datasets are available? 

- Which species are most represented? 

**Hint:** The `mcols()` function will extract metadata associated with each
entry in the `AnnotationHub`.

::: {.callout-hint .icon .callout-note}

## Hint

Create the hub with `AnnotationHub::AnnotationHub()`, extract its metadata
`data.frame` with `mcols()`, then `table()` the `species` (or `sourcetype`)
column, `sort()` it and look at the `tail()`.

```{r}
ah <- AnnotationHub::AnnotationHub()
ah_df <- AnnotationHub::mcols(ah)
table(ah_df$______) |> sort() |> tail(10)
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
ah <- AnnotationHub::AnnotationHub()
ah_df <- AnnotationHub::mcols(ah)
table(ah_df$species) |> sort() |> tail(10)
table(ah_df$sourcetype) |> sort() |> tail(10)
```

:::

:::

The Epigenome roadmap project by the Broad Institute performed H3K4me3 ChIP-seq
in human (among other marks). They generated a library labelled 'LL227' 
([GEO ID: GSM409308](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM409308)).

::: {.callout-question .icon .callout-note}

Can you query the AnnotationHub for this library?

::: {.callout-hint .icon .callout-note}

## Hint

`AnnotationHub::query(ah, "<pattern>")` searches all metadata fields for a
pattern. Try the library label mentioned above.

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
AnnotationHub::query(ah, 'LL227')
```

:::

:::

::: {.callout-question .icon .callout-note}

- Which signal corresponds to a ChIP-seq coverage track? How can you download the associated file?

::: {.callout-hint .icon .callout-note}

## Hint

Among the query hits, look for a `BigWigFile` resource (the coverage track).
Extracting an entry with `ah[["<AH id>"]]` downloads it (and caches it locally).

```{r}
bw <- ah[['______']]
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
bw <- ah[['AH34904']]
```

:::

:::

::: {.callout-question .icon .callout-note}

What is the path of the file now stored on disk? How can you load the file in R?

::: {.callout-hint .icon .callout-note}

## Hint

`BiocIO::resource()` returns the local path of the cached file, and
`BiocIO::import()` reads it into R.

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
H3K4me3_bw <- BiocIO::resource(bw)
cov <- BiocIO::import(bw)
```

:::

:::

## Create a `AggregatedCoverage` object

This `AggregatedCoverage` should contain aggregated H3K4me3 coverage over the 
TSSs of all **forward** human genes. 

::: {.callout-question .icon .callout-note}

Three arguments are required: 

- `bw_file`: the path to a disk-stored coverage track file (e.g. the path to a `bw` file)
- `features_file`: the path to a disk-stored features file (e.g. the path to a `bed` file)
- `width`: an integer specifying the width to use for each genomic locus

::: {.callout-hint .icon .callout-note}

## Hint

- `bw_file`: the local path returned by `BiocIO::resource(bw)`.
- `features_file`: derive TSSs of **forward** genes from the `TxDb`
  (`genes()`, subset `strand == "+"`, `resize(width = 1, fix = "start")`), then
  `rtracklayer::export()` them to a `bed` file — `AggregatedCoverage()` expects
  a *file path*, not a `GRanges`.
- `width`: an integer, e.g. `8000L` for a ±4 kb window.

```{r}
forward_genes <- genes[strand(genes) == '+']
forward_tss <- resize(forward_genes, width = 1, fix = 'start')
export(forward_tss, "forward_tss.bed")
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

- Preparing `bw_file` argument:

```{r}
bw_file <- H3K4me3_bw
```

- Preparing `features_file` argument: 

```{r}
library(GenomicRanges)
library(rtracklayer)
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene::TxDb.Hsapiens.UCSC.hg19.knownGene
genes <- GenomicFeatures::genes(txdb)
forward_genes <- genes[strand(genes) == '+']
forward_tss <- resize(forward_genes, width = 1, fix = 'start')
export(forward_tss, "forward_tss.bed")
features_file <- 'forward_tss.bed'
```

- Preparing `width` argument: 

```{r}
width <- 8000L
```

:::

:::

Now that all the arguments are ready, one can create an `AggregatedCoverage` 
object. 

::: {.callout-question .icon .callout-note}

Create a `AggregatedCoverage` object summarizing the coverage
of H3K4me3 over the TSSs (± 4kb) of all forward human genes.

::: {.callout-hint .icon .callout-note}

## Hint

This is the very constructor you wrote in Day 4 — call it with the three
arguments you just prepared.

```{r}
AC <- AggregatedCoverage(bw_file, features_file, width = width)
```

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
AC <- AggregatedCoverage(bw_file, features_file, width = width)
AC
```

:::

:::

::: {.callout-question .icon .callout-note}

Plot the resulting `AC` object.

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
plot(AC)
```

:::

:::

## Add a vignette to your package

The previous section reuses publicly available datasets and 
resources provided by Bioconductor to illustrate a use case for our 
package. 

::: {.callout-question .icon .callout-note}

- Initiate a vignette using `biocthis` template; 

- Fill out the vignette header with relevant information (namely, the `author` field);

- Describe the use case developed in the previous section.

::: {.callout-hint .icon .callout-note}

## Hint

`biocthis::use_bioc_vignette()` scaffolds a `Bioconductor`-style vignette under
`vignettes/`. It takes the vignette `name` and a `title`.

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
biocthis::use_bioc_vignette(name = "JacquesTestPackage", title = "Introduction to my package")
```

:::

:::

::: {.callout-question .icon .callout-note}

Because the vignette relies on several packages, these need to be declared in
the DESCRIPTION file. Make sure to update it.

::: {.callout-hint .icon .callout-note}

## Hint

`usethis::use_package("<pkg>")` adds a package to the `Imports:` field of your
DESCRIPTION (use `type = "Suggests"` for packages only needed by the vignette or
tests). Call it once per dependency.

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
usethis::use_package("BiocStyle")
usethis::use_package("TxDb.Hsapiens.UCSC.hg19.knownGene")
usethis::use_package("GenomicFeatures")
usethis::use_package("GenomicRanges")
usethis::use_package("AnnotationHub")
usethis::use_package("BiocIO")
```

:::

:::


## Create pkgdown website and host it on GitHub

::: {.callout-question .icon .callout-note}

Check `pkgdown` documentation. What is the recommended way to add pkgdown support
for your package?

::: {.callout-hint .icon .callout-note}

## Hint

`usethis::use_pkgdown()` sets up everything pkgdown needs (a `_pkgdown.yml`
config, and the relevant `.gitignore`/`.Rbuildignore` entries).

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
usethis::use_pkgdown()
```

:::

:::

::: {.callout-question .icon .callout-note}

Can you now build your package? What does the building step do exactly?

::: {.callout-hint .icon .callout-note}

## Hint

`pkgdown::build_site()` renders the whole documentation website (home page from
the README, function reference from the `man/` pages, and articles from the
vignettes) into the `docs/` folder.

:::

::: {.callout-answer .icon .callout-note collapse=true}

```{r}
pkgdown::build_site()
```

:::

:::

::: {.callout-question .icon .callout-note}

- Comment on why setting up a pkgdown website helps with checking that your package reliably works? 

- Commit your changes. Did that include your newly created `docs` folder? Why? 

- Remove `docs` from `.gitignore`, then commit the `docs/` folder with git. 

- Now push all your commits to your github remote repository, navigate to its webpage, and enable `Pages` deployment from the `docs` folder. Check the resulting website. 

:::
