---
title: "Ex. 03-3: Integrating ChIP-seq results together"
output: 
    rmdformats::readthedown: 
        highlight: tango
        preserve_yaml: true
        df_print: tibble
        toc_depth: 4
        css: ../../../custom.css
---

## 1. Aggregating all datasets together

Let's load all the peaks in R, using `rtracklayer` package.

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r eval = FALSE}
TFs <- list.files(pattern = '*.narrowPeak$', path = '~/Share/day03/nf-chip-results/bwa/mergedLibrary/macs/narrowPeak')
TFs <- unique(gsub('_.*', '', TFs))
require(rtracklayer)
list_TF_peaks <- lapply(TFs, function(TF) {
    TF_files <- list.files(pattern = paste0(TF, '.*narrowPeak$'), path = '~/Share/day03/nf-chip-results/bwa/mergedLibrary/macs/narrowPeak', full.names = TRUE)
    peaks <- GenomicRanges::reduce(do.call(c, lapply(TF_files, rtracklayer::import)))
    return(peaks)
})
names(list_TF_peaks) <- TFs
all_peaks <- GenomicRanges::reduce(do.call(c, lapply(
    list.files(pattern = '.*narrowPeak$', path = '~/Share/day03/nf-chip-results/bwa/mergedLibrary/macs/narrowPeak', full.names = TRUE), rtracklayer::import
)))
```
</p></details><br>

> Is there redundancy in TF binding? 

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r eval = FALSE}
length(all_peaks)
sum(lengths(list_TF_peaks))
```
</p></details><br>

> How can we summarize all the TF binding events in a single data.frame? 

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r eval = FALSE}
`%over%` <- IRanges::`%over%`
df <- lapply(names(list_TF_peaks), function(TF) {
    d <- data.frame(
        TF = all_peaks %over% IRanges::subsetByOverlaps(all_peaks, list_TF_peaks[[TF]])
    )
    colnames(d) <- TF
    return(d)
}) 
df <- do.call(cbind, df)
#
dim(df)
colSums(df)
table(rowSums(df))
```
</p></details><br>

## 2. Calculating an enrichment of binding between pairs of TFs

> Calculate the odds ratio between (1) binding of Hog1 and (2) binding of Hot1, over all the bound peaks  

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r eval = FALSE}
TF1 <-'Hot1'
TF2 <-'Hog1'
peaks_with_both <- sum(df[, 'Hot1'] & df[, 'Hog1'])
peaks_with_1 <- sum(df[, 'Hot1'] & !df[, 'Hog1'])
peaks_with_2 <- sum(!df[, 'Hot1'] & df[, 'Hog1'])
peaks_with_none <- sum(!df[, 'Hot1'] & !df[, 'Hog1'])
mat <- matrix(c(peaks_with_both, peaks_with_1, peaks_with_2, peaks_with_none), ncol = 2)
fisher.test(mat)
```
</p></details><br>

> Generalize it for all TFs  

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r eval = FALSE}
d <- lapply(TFs,  function(TF1) {
    df <- data.frame(lapply(TFs, function(TF2) {
        if (TF1 == TF2) {
            NA
        } else {
            peaks_with_both <- sum(df[, TF1] & df[, TF2])
            peaks_with_1 <- sum(df[, TF1] & !df[, TF2])
            peaks_with_2 <- sum(!df[, TF1] & df[, TF2])
            peaks_with_none <- sum(!df[, TF1] & !df[, TF2])
            mat <- matrix(c(peaks_with_both, peaks_with_1, peaks_with_2, peaks_with_none), ncol = 2)
            fisher.test(mat)$estimate
        }
    }))
    colnames(df) <- TFs
    rownames(df) <- TF1
    return(df)
}) 
d <- do.call(rbind, d)
```
</p></details><br>

> Plot results as a heatmap  

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r eval = FALSE}
library(tidyverse)
dd <- d %>% 
    rownames_to_column('TF1') %>% 
    gather('TF2', 'odds_ratio', -TF1)
p <- ggplot(dd, aes(x = TF1, y = TF2, fill = odds_ratio)) + 
    geom_tile() + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + 
    scale_fill_gradientn(colours = c('white', 'orange', 'darkred'))
p
# Clamp values to a certain range
q_max <- quantile(dd$odds_ratio, 0.95, na.rm = TRUE)
dd$odds_ratio[dd$odds_ratio > q_max] <- q_max
# Order rows
d[is.na(d)] <- 0
o <- hclust(dist(d))$order
o <- hclust(dist(t(df)))$order
dd %<>% mutate(TF1 = factor(TF1, levels = TFs[o])) %>% 
    mutate(TF2 = factor(TF2, levels = TFs[o]))
# Re-plot
p <- ggplot(dd, aes(x = TF1, y = TF2, fill = odds_ratio)) + 
    geom_tile() + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + 
    scale_fill_gradientn(colours = c('white', 'orange', 'darkred'))
p
```
</p></details><br>

> Plot results as a graph  

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r eval = FALSE}
d_counts <- lapply(TFs,  function(TF1) {
    vec <- data.frame(lapply(TFs, function(TF2) {
        ifelse(
            which(TFs == TF2) <= which(TFs == TF1), 
            0, 
            sum(df[, TF1] & df[, TF2]) / sum(df[, TF2] | df[, TF1])
        )
    }))
    names(vec) <- TFs
    return(vec)
}) 
d_counts <- do.call(rbind, d_counts)
row.names(d_counts) <- TFs
#
dd <- d_counts %>% 
    rownames_to_column('TF1') %>% 
    gather('TF2', 'weight', -TF1) %>% 
    # mutate(weight = ifelse(weight < 0.1, 0, weight)) %>% 
    filter(weight > 0.15 & weight < 1) %>% 
    filter(TF1 != 'Rsc9-TAP' & TF2 != 'Rsc9-TAP')
gr <- igraph::graph_from_data_frame(dd, directed = FALSE)
plot(gr)
```
</p></details><br>

## Conclusions

> Can you speculate on how the transcription factors combine on the chromatin? 

> What was your initial hypothesis regarding the different types of peak sets (Ex. 03-1)? Does this match your observations here? 